---
title: "phylogenetic examples"
author: "Christof Neumann"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: true
    keep_tex: false
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{phylogenetic_examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
csl: apa.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
suppressPackageStartupMessages(library(EloSteepness))
suppressPackageStartupMessages(library(EloSteepness.data))
suppressPackageStartupMessages(library(rstan))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(brms))
suppressPackageStartupMessages(library(xtable))
# suppressPackageStartupMessages(library(kableExtra))
```


This document contains the code to replicate the two analyses of macaque hierarchy steepness (one analysis in the main text, one in the appendix) from *Extending Bayesian Elo-rating to quantify the steepness of dominance hierarchies* (https://doi.org/10.1101/2022.01.28.478016).
It was built using `EloSteepness` (version `r packageVersion("EloSteepness")`) and `EloSteepness.data` (version `r  packageVersion("EloSteepness.data")`).



```{r setup}
library(EloSteepness)
library(EloSteepness.data)
library(rstan)
library(cmdstanr)
library(ape)
library(brms)
library(xtable) # just for table generation in this document
# library(kableExtra) # just for table generation in this document
```

# Comparative example from the main text

```{r}
# path to raw model code
mod1 <- system.file("extdata/phylogenetic_examples/comparative_steepness_model_1.stan",
                    package = "EloSteepness.data")
file.exists(mod1)
```

```{r}
# load tree, matrices and key table
data("Primates301_nex")
data("phylo_matrices")
data("phylo_key")

# prune tree to available species
species <- unique(phylo_key$species_fixed)
tree <- keep.tip(Primates301_nex, species)
# plot(tree)
length(tree$tip.label) == length(species) # sanity check

# sort matrices and key so they have the same order
interaction_matrices <- phylo_matrices[sort(names(phylo_matrices))]
key <- phylo_key[order(phylo_key$datasetname), ]
all(names(interaction_matrices) == key$datasetname) # sanity check

# summary data for each group
specdata <- data.frame(datasetname = key$datasetname,
                       spec = gsub(" ", "_", key$species_fixed),
                       groupsize = NA,
                       n_interactions = NA,
                       dyads = NA,
                       prunks = NA)

for (m in seq_len(nrow(specdata))) {
  mat <- interaction_matrices[[specdata$datasetname[m]]]
  specdata$groupsize[m] <- ncol(mat)
  specdata$n_interactions[m] <- sum(mat)
  specdata$dyads[m] <- ncol(mat) * (ncol(mat) - 1) / 2
  specdata$prunks[m] <- EloRating::prunk(mat)[1]
}
```

For the sake of this vignette we reduce the number of species and also the number of interactions in the networks.
We do this just because it speeds up the sampling from the model.
With the full data set, this would take several hours (about four on my fairly old laptop).
With the reduced data set this takes less time to run (but still takes a few minutes).

```{r}
# ignore these lines if you want to run the full data set
specdata <- specdata[specdata$groupsize < 7, ]
interaction_matrices <- interaction_matrices[specdata$datasetname]
tree <- keep.tip(tree, specdata$spec)
interaction_matrices <- lapply(interaction_matrices, function(x) ceiling(x / 5))
```


```{r}
# prepare the data set as list for sending it to Stan
xdata <- list()
xdata$n_datasets <- length(interaction_matrices)
xdata$n_ids_per_dataset <- as.integer(unlist(lapply(interaction_matrices, ncol)))
xdata$n_total_ids <- sum(xdata$n_ids_per_dataset)
xdata$n_interactions_per_dataset <- as.integer(unlist(lapply(interaction_matrices, sum)))
xdata$n_total_interactions <- sum(xdata$n_interactions_per_dataset)
xdata$interaction_index <- as.integer(seq_len(xdata$n_total_interactions))
xdata$individual_index <- as.integer(seq_len(xdata$n_total_ids))


# generate sequences
set.seed(1)
s <- lapply(interaction_matrices, randomized_sequence_from_matrix)
# global winner and loser indices
xdata$winner <- do.call("rbind", lapply(s, function(x)x$winnermat))
xdata$loser <- do.call("rbind", lapply(s, function(x)x$losermat))

xdata$index_dataset_interactions_start <- cumsum(xdata$n_interactions_per_dataset) -
                                            xdata$n_interactions_per_dataset + 1
xdata$index_individuals_start <- cumsum(xdata$n_ids_per_dataset) -
                                   xdata$n_ids_per_dataset + 1

xdata$y <- rep(1, xdata$n_total_interactions)

# phylogeny-related
n_spec <- length(tree$tip.label)
tree_mat <- vcv.phylo(tree, corr = TRUE)
species_index <- integer(nrow(specdata))
for (i in seq_len(nrow(specdata))) {
  species_index[i] <- which(colnames(tree_mat) == specdata$spec[i])
}

# for phylogeny part of the model
xdata$N_phyl <- n_spec # N_1
xdata$N_coeff_phyl <- 1 # number of parameters: 1 = SD for phyl 'intercepts'
xdata$spec_index <- species_index
xdata$chol_mat <- t(chol(tree_mat)) # Cholesky factor of phylogenetic correlation matrix
xdata$phyl_predictor <- rep(1, nrow(specdata)) # group-level predictor values

# for repeated measures part of the model
xdata$N_repeated <- n_spec # the same as for the phylogeny part
# number of parameters: 1 = SD for repeated measurements (species intercepts)
xdata$N_coeff_repeated <- 1
xdata$index_rep_measures <- species_index # the same as for the phyl part
# group-level predictor values, the same as for the phyl part
xdata$rep_measures_predictor <- rep(1, nrow(specdata))
```


```{r modelcompile1, cache=FALSE, warning = FALSE, include=FALSE, results = 'hide'}
# compile stan model
xmod1 <- cmdstan_model(stan_file = mod1, compile = TRUE)
```

```{r modelcompile1_display, eval = FALSE}
# compile stan model
xmod1 <- cmdstan_model(stan_file = mod1, compile = TRUE)
```


```{r modelfit1, cache=FALSE, results='hide', echo=FALSE, warning = FALSE, include=FALSE}
# sampling
res1 <- suppressMessages(xmod1$sample(data = xdata,
                                      refresh = 0,
                                      parallel_chains = 4,
                                      iter_warmup = 1000,
                                      iter_sampling = 500,
                                      seed = 1,
                                      adapt_delta = 0.8))
```

```{r modelfit1_demo, eval=FALSE}
# sampling
res1 <- xmod1$sample(data = xdata,
                     refresh = 100,
                     parallel_chains = 4,
                     iter_warmup = 1500,
                     iter_sampling = 2500,
                     seed = 1,
                     adapt_delta = 0.9)
```


```{r, cache=FALSE}
stanfit <- rstan::read_stan_csv(res1$output_files())
xsummary <- summary(stanfit)$summary
round(summary(stanfit, pars = c("sd_repeated_measures", "sd_phyl"))$summary, 2)

# optional saving of results
# save(xmod1, xsummary, res1, stanfit, xdata, specdata, tree, file = "comp_analysis_1.RData")
```


```{r, out.width="100%", fig.width=9, fig.height=4, fig.retina=2}
par(mfrow = c(1, 2), family = "serif")
par(family = "serif", mgp = c(1.8, 0.7, 0), las = 1, mar = c(2, 1, 1, 1))

plot.phylo(tree, cex = 0.8, show.tip.label = FALSE, x.lim = c(0, 19))
tips <- tree$tip.label
for (i in seq_along(tips)) {
  x <- sum(specdata$spec == tips[i])
  y <- gsub("_", " ", tips[i])
  tiplabels(text = bquote(italic(.(y))~.(paste0(" (", x, ")"))),
            tip = i, bg = NULL, adj = 0, frame = "none", offset = 0.3, cex = 0.8)
}

par(mar = c(3, 2.5, 1.5, 1))

# posteriors
x <- as.numeric(extract(stanfit, pars = c("sd_phyl"))$sd_phyl)
y <- as.numeric(extract(stanfit, pars = c("sd_repeated_measures"))$sd_repeated_measures)
# priors
xvals <- rstudent_t(10000, 3, 0, 2.5)
while (any(xvals < 0)) {
  xvals[xvals < 0] <- brms::rstudent_t(sum(xvals < 0), 3, 0, 2.5)
}
p3 <- density(xvals, adjust = 1)

p1 <- density(x, adjust = 1)
p2 <- density(y, adjust = 1)
yr <- c(0, max(p1$y, p2$y) * 1.05)
xr <- c(0, max(p1$x, p2$x) * 1.05)
xr[2] <- 2
plot(0, 0, type = "n", xlim = xr, ylim = yr, xaxs = "i", yaxs = "i", las = 1,
     axes = FALSE, xlab = "SD", ylab = "density")
axis(1)

polygon(c(p1$x, rev(p1$x)), c(rep(0, length(p1$y)), rev(p1$y)),
        border = NA, col = adjustcolor("#3B99B1", 0.5))
polygon(c(p2$x, rev(p2$x)), c(rep(0, length(p2$y)), rev(p2$y)),
        border = NA, col = adjustcolor("#EACB2B", 0.5))
polygon(c(p3$x, rev(p3$x)), c(rep(0, length(p3$y)), rev(p3$y)),
        border = NA, col = adjustcolor("grey", 0.7))
points(p1$x, p1$y, type = "l", lwd = 0.5)
points(p2$x, p2$y, type = "l", lwd = 0.5)
points(p3$x, p3$y, type = "l", lwd = 0.5)
box(bty = "l")

legend("topright", ncol = 1, col = c("#3B99B1", "#EACB2B", "grey"),
       legend = c("phylogenetic", "repeated measurements", "prior for both"),
       pch = 15, bty = "n")
```


\newpage

# Reanalysis of Balasubramaniam et al data set

```{r}
# path to raw model code
mod2 <- system.file("extdata/phylogenetic_examples/comparative_steepness_model_2.stan",
                    package = "EloSteepness.data")
file.exists(mod2)
```


```{r}
# tree data
data("Primates301_nex")

# data from Balasubramaniam et al 2012 (table 2)
sp <- c("Macaca_assamensis", "Macaca_fascicularis", "Macaca_fascicularis",
        "Macaca_fuscata", "Macaca_fuscata", "Macaca_mulatta", "Macaca_mulatta",
        "Macaca_nigra", "Macaca_radiata", "Macaca_sylvanus", "Macaca_thibetana",
        "Macaca_thibetana", "Macaca_tonkeana", "Macaca_tonkeana")
st <- c(0.65, 0.94, 0.79, 0.56, 0.92, 0.65, 0.78,
        0.49, 0.60, 0.45, 0.87, 0.80, 0.22, 0.20)
specdata <- data.frame(spec = sp,
                       steepness = st)

# prune tree to 9 macaque species
species <- unique(specdata$spec)
tree <- keep.tip(Primates301_nex, species)
# plot(tree)
length(tree$tip.label) == length(species) # sanity check


# prepare data set to be handed over to stan
xdata <- list()
xdata$N <- nrow(specdata)
xdata$the_steepness <- specdata$steepness


# phylogeny-related
tree_mat <- vcv.phylo(tree, corr = TRUE) # covariance matrix
n_spec <- length(tree$tip.label)
# index for species
species_index <- integer(nrow(specdata))
for (i in seq_len(nrow(specdata))) {
  species_index[i] <- which(colnames(tree_mat) == specdata$spec[i])
}

# for phylogeny part
xdata$N_phyl <- n_spec # number of levels
xdata$N_coeff_phyl <- 1 # number of pars: SD for phyl 'intercepts'
xdata$spec_index <- species_index
xdata$chol_mat <- t(chol(tree_mat)) # cholesky factor of phylogenetic correlation matrix
xdata$phyl_predictor <- rep(1, nrow(specdata)) # group-level predictor values

# for repeated measures part
xdata$N_repeated <- n_spec # the same as for the phyl part (number of levels)
# number of pars: SD for repeated measurements (species intercepts)
xdata$N_coeff_repeated <- 1
xdata$index_rep_measures <- species_index # the same as for the phyl part
# group-level predictor values, the same as for the phyl part
xdata$rep_measures_predictor <- rep(1, nrow(specdata))
```

```{r modelcompile2, cache=FALSE, warning = FALSE, include=FALSE, results='hide'}
# compile stan model
xmod2 <- cmdstan_model(stan_file = mod2, compile = TRUE)
```


```{r modelcompile2_display, eval=FALSE}
# compile stan model
xmod2 <- cmdstan_model(stan_file = mod2, compile = TRUE)
```

```{r modelfit2, cache=FALSE, results='hide', echo=FALSE, warning = FALSE}
# sampling
res2 <- suppressMessages(xmod2$sample(data = xdata,
                   refresh = 0,
                   iter_warmup = 1000,
                   iter_sampling = 500,
                   parallel_chains = 4,
                   seed = 123,
                   adapt_delta = 0.8))
```


```{r modelfit2_demo, eval=FALSE}
# sampling
res2 <- xmod2$sample(data = xdata,
                     refresh = 0,
                     parallel_chains = 4,
                     seed = 123,
                     adapt_delta = 0.9)
```







```{r, out.width="100%", fig.width=9, fig.height=4, fig.retina=2}
# convert back to rstan (for easier handling of output)
stanfit2 <- read_stan_csv(res2$output_files())
xsummary2 <- summary(stanfit2)$summary
round(summary(stanfit2,
      pars = c("sd_phyl", "sd_repeated_measures", "phi", "Intercept"))$summary, 2)

# optional saving of results
# save(xmod2, xsummary2, res2, stanfit2, xdata, specdata, tree, file = "comp_analysis_2.RData")

# plot in the manuscript
sd_phyl <- sprintf("%.2f", xsummary2["sd_phyl[1]", 1])
sd_rep <- sprintf("%.2f", xsummary2["sd_repeated_measures[1]", 1])

par(mfrow = c(1, 2), family = "serif")
par(family = "serif", mgp = c(1.8, 0.7, 0), las = 1, mar = c(2, 1, 1, 1))

# plot tree with number of groups
ape::plot.phylo(tree, cex = 0.8, show.tip.label = FALSE, x.lim = c(0, 19))
tips <- tree$tip.label
for (i in seq_along(tips)) {
  x <- sum(specdata$spec == tips[i])
  y <- gsub("_", " ", tips[i])
  ape::tiplabels(text = bquote(italic(.(y))~.(paste0(" (", x, ")"))),
                 tip = i, bg = NULL, adj = 0,
                 frame = "none", offset = 0.3, cex = 0.8)
}
par(mar = c(3, 2.5, 1.5, 1))

# extract posterior samples for SDs and corresponding priors
est_phyl <- as.numeric(extract(stanfit2, pars = c("sd_phyl"))$sd_phyl)
est_rep <- as.numeric(extract(stanfit2, pars = c("sd_repeated_measures"))$sd_repeated_measures)

pr_phyl <- as.numeric(extract(stanfit2, pars = c("prior_sd_phyl"))$prior_sd_phyl)
pr_rep <- as.numeric(extract(stanfit2, pars = c("prior_sd_repeated_measures"))$prior_sd_repeated_measures)

# and densities of posteriors
est_phyl <- density(est_phyl, adjust = 1)
est_rep <- density(est_rep, adjust = 1)

# and for priors
pr_phyl <- density(pr_phyl[pr_phyl < 10], adjust = 1) # truncated for smoother display
pr_rep <- density(pr_rep[pr_rep < 10], adjust = 1)


# data ranges for plot
yr <- c(0, max(est_phyl$y, est_rep$y) * 1.05)
xr <- c(0, max(est_phyl$x, est_rep$x) * 1.05)


plot(0, 0, type = "n", xlim = xr, ylim = yr, xaxs = "i", yaxs = "i",
     las = 1, axes = FALSE, xlab = "SD", ylab = "density")
axis(1)

polygon(c(est_phyl$x, rev(est_phyl$x)), c(rep(0, length(est_phyl$y)), rev(est_phyl$y)),
        border = NA, col = adjustcolor("#3B99B1", 0.5))
polygon(c(est_rep$x, rev(est_rep$x)), c(rep(0, length(est_rep$y)), rev(est_rep$y)),
        border = NA, col = adjustcolor("#EACB2B", 0.5))
polygon(c(pr_phyl$x, rev(pr_phyl$x)), c(rep(0, length(pr_phyl$y)), rev(pr_phyl$y)),
        border = NA, col = adjustcolor("grey", 0.4))
polygon(c(pr_rep$x, rev(pr_rep$x)), c(rep(0, length(pr_rep$y)), rev(pr_rep$y)),
        border = NA, col = adjustcolor("grey", 0.4))

points(est_phyl$x, est_phyl$y, type = "l", lwd = 0.5)
points(est_rep$x, est_rep$y, type = "l", lwd = 0.5)
points(pr_phyl$x, pr_phyl$y, type = "l", lwd = 0.5)
points(pr_rep$x, pr_rep$y, type = "l", lwd = 0.5)

box(bty = "l")

legend("topright", ncol = 1, col = c("#3B99B1", "#EACB2B", "grey"),
       legend = c("phylogenetic", "repeated measurements", "priors"),
       pch = 15, bty = "n")
```

\newpage


# Data summary

Table \ref{tab:domnets} provides data summaries and references for the analyses run above.

```{r, echo=FALSE, results='asis'}
data("phylo_key")
data("phylo_matrices")
m <- phylo_key
m$species_fixed[is.na(m$species_fixed)] <- ""
m <- m[order(m$datasetname), ]
p <- phylo_matrices[m$datasetname]
m$group_size <- unlist(lapply(p, ncol))
m$interactions <- unlist(lapply(p, sum))
m$sparseness <- round(unlist(lapply(p, function(x)prunk(x)[1])), 2)

m <- m[, c("datasetname", "species_fixed", "common_name", "group_size", "interactions", "sparseness", "reference")]
m$species_fixed <- gsub("_", " ", m$species_fixed)

# create reference keys that work inside tables...
x <- unique(unlist(regmatches(x = m$reference, gregexpr("(@[a-z-]+[0-9]{4}[a-d]{0,1})", text = m$reference))))
cat(paste(paste0("(ref:", gsub(pattern = "^@", "", x), ") ", x), collapse = "\n\n"))



# saves <- c()
# for (i in seq_len(length(m$reference))) {
#   rkey <- gsub("@", "", m$reference[i])
#   rkey <- gsub(" ", "", rkey)
#   rkey <- gsub(";", ",", rkey)
#   saves <- c(saves, rkey)
# }
# saves[saves == "Micheletta(pers.comm.)" ] <- "Micheletta (pers. comm.)" 
```

\setlength\LTleft{-0.8in}

\setlength\LTright{-1in}

```{r, echo=FALSE, results='asis'}
for (i in 1:nrow(m)) {
  x <- m$reference[i]
  y <- unlist(regmatches(x = x, gregexpr("(@[a-z-]+[0-9]{4}[a-d]{0,1})", text = x)))
  for (k in y) {
    x <- gsub(pattern = k, paste0("(ref:", gsub(pattern = "^@", "", k), ")"), x)
  }
  m$reference[i] <- x
}

m$datasetname <- gsub("_", "\\\\_", m$datasetname)
m$species_fixed <- paste0("\\textit{", m$species_fixed, "}")

colnames(m) <- gsub("\\_", " ", colnames(m))
colnames(m)[1:2] <- c("data set", "species")
# colnames(m)

d <- c("d", "s", "s", "s", "d", "d", "f", "s")

xtab <- xtable(m[, 1:7], 
               label = "tab:domnets", 
               caption = "Macaque dominance interaction networks", 
               display = d)
align(xtab) <- c("p{0.1\\textwidth}", 
                 "p{0.23\\textwidth}", 
                 "p{0.19\\textwidth}",
                 "p{0.2\\textwidth}",
                 "p{0.04\\textwidth}",
                 "p{0.08\\textwidth}", 
                 "p{0.08\\textwidth}",
                 "p{0.22\\textwidth}")

addtorow <- list()
addtorow$pos <- list()
addtorow$pos[[1]] <- c(0)
addtorow$command <- c(paste(
  "\\hline \n",
  "\\endhead \n",
  "\\hline \n",
  "{\\footnotesize Continued on next page} \n",
  "\\endfoot \n",
  "\\endlastfoot \n",
  sep = ""))

# addtorow <- list(pos = list(0),command = "\\hline \\endhead ")
print(xtab, 
      tabular.environment = "longtable",
      caption.placement = "top", 
      sanitize.text.function = function(x) {x}, 
      booktabs = TRUE,
      floating = FALSE,
      comment = FALSE,
      include.rownames = FALSE, 
      size = "\\footnotesize", #  size="\\fontsize{7pt}{9pt}\\selectfont", 
      hline.after = c(-1, nrow(xtab)),
      add.to.row = addtorow)
```


```{r, echo=FALSE, eval=FALSE, results='markup'}
m$reference <- saves
x <- m[, -3]
x$datasetname <- gsub("_", "\\\\_", x$datasetname)
x$reference[saves != "Micheletta (pers. comm.)"] <- paste0("\\cite{", x$reference[saves != "Micheletta (pers. comm.)" ], "}")

kable(x, format = "latex", 
      longtable = TRUE,
      caption = "Macaque dominance interaction networks",
      row.names = FALSE, 
      booktabs = TRUE,
      linesep = "", 
      escape = FALSE,
      col.names = c("data set", "scientific name", "group size", "interactions", "sparseness", "source")) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header"), 
                repeat_header_text = "(continued)", 
                full_width = TRUE) %>% 
  column_spec(column = 1, width = "1.7in") %>%
  column_spec(column = 2, width = "1.1in", italic = TRUE) %>%
  column_spec(column = 3, width = "0.4in") %>%
  column_spec(column = 4, width = "0.4in") %>%
  column_spec(column = 5, width = "0.7in") %>%
  column_spec(column = 6, width = "1.6in")

```


\newpage


# References
